Palomar 5 Gaps

### 2024-10-21 paper/main.tex
the bib should work now

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{montuori07}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{keenan75}. These predictions occured well before the first detections which were made by \citet{carl_j_grillmair_globular_1995} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{ibata1994} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{lehmann97, testa00, leon00, siegel01, lee03}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,odenkirchen02, odenkirchen03} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{grillmair06a, grillmair06b, belokurov06, grillmair09}. The review of Bonaca properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{prousti16}.

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{mateu23} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose memebers have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020 reports that the Phenoix stream is dynamically consistent from a globular cluster \footnote{see bovy for a description on the difference between dwarf and cluster streams} yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times. Nicolas Martin in 2022 showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, Vitral and Bouldrini demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are over XXX cluster in the internal regional of the galaxy, a region overwhich stream finder was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, Balbinot demonstrates that internal dynamics preferentially kicks out small stars, which have lower luminosities and can make detection more difficult.

  One of the interests in studying stellar streams, is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. For example, \citet{varghese2011stellar} demonstrated how stellar streams can infer the mass and scale parameters of \paola{dark matter} haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{bonaca14} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams that yield the most information about the galactic potential. However, the use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{law10} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{helmi04, johnston05, law05}; \citet{bovy16} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. Nonetheless, a recent goal of stellar stream astronomy was realized by \citet{ibata24} constructed the most accurate model of the Milky Way to date based on a set of about 90 streams.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by \citet{vegetti12}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{amorisco21}]{green05, wang20}.

  \citet{rodrigo_ibata_uncovering_2002} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{r_g_carlberg_pal_2012} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. Carlberg 2013 further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{bonaca_spur_2018} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship from M et al 20XX.

  Our study focuses on the influence of globular clusters on stellar streams, specifically examining whether globular clusters could produce gaps similar to those expected from dark matter subhalos. If globular clusters do indeed create such gaps, this factor must be considered when using stellar streams to detect dark matter. An additional and unique aspect of our study is that we focus on modeling real stellar streams and not just pure simulated streams. We focus on the Palomar 5 stream, which is well-studied, has observable gaps (bonaca et al 20XX), and is highly visible due to its position above the galactic disk. We present a result from one of our simulations in Fig.~\ref{fig:stream_on_sky}, and the following sections describe our modeling of the stream, the gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations affecting the stream.

@@

      \citet{sanders2016dynamics} extended the analysis of Erkal et al. (2015), demonstrating that action-angle variables provide a useful coordinate system for analyzing stream evolution, as actions are conserved and the angles associated with the stream's growth evolve linearly over time. Although we became aware of this work only after completing our analysis, we note that $\tau$ is a suitable approximation and behaves similarly to the angle variable corresponding to the azimuthal action: $\tau \approx \theta_{\phi,i} - \theta_{\phi,\text{GC}}$.

>>>>

  Stellar streams, several-kpc long structures formed by the tidal disruption of globular clusters or dwarf galaxies orbiting a host galaxy. These tidal forces arise due to differential gravitational pulls across extended objects, causing stars farther from the galactic center to lag behind while those closer are pulled away. This stretching creates two tidal tails that trace the cluster's orbit, unless in the closest vicinity to the object \citep{2007ApJ...659.1212M}. Numerical predictions of this phenomenon exist since the 1970s \citep[see, for example][]{1975AJ.....80..290K}. These predictions occured well before the first detections which were made by \citet{1995AJ....109.2553G} around some Galactic globular clusters. Interestingly, this work was a nearly contemporary discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way. Subsequent studies\footnote{\citep[see][for more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius.]{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster, which were shown to be extended for more than $10^\circ$ degrees in the sky. These early works stimulated a vigorous search for stellar streams in the Galaxy in the following years \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2006ApJ...637L..29B, 2009ApJ...693.1118G}. The review of Bonaca properly asserts that we have envolved into a new area of stellar stream astronomy Gaia astrometric mission \citep{2016A&A...595A...1G}.

  Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and spectroscopy. The development and deployment of the Streamfinder algorithm has led to the detection of about 90 stellar streams in the Milky Way. \citet{2023MNRAS.520.5225M} provides a summary of these streams and tabulates that about 20 are associated to modern day globular clusters. Thus, currently there are about 70 progenitor-less streams and about 145 stream-less clusteres. The solo-streams can be explained by a larger globular cluster population whose memebers have been disolved due to two body relaxation and tidal stripping. For instance, Z Wan 2020 reports that the Phenoix stream is dynamically consistent from a globular cluster \footnote{see bovy for a description on the difference between dwarf and cluster streams} yet far more metal poor than any cluster ever observed, galactic or otherwise. This implies that the stream is from a completely disolved globular cluster which comes from a population of clusters that have not survived until modern times. Nicolas Martin in 2022 showed that many very metal poor streams exist in the Milky Way. On the other hand, it is possible that some globular clusters do not loose a significant number of stars to create tidal tails. For instance, Vitral and Bouldrini demonstrated that clusters existing in dark matter subhaloes can place the tidal radius far beyond their half mass radii and sheild stars from escape. Another factor is indeed detectability. Additionally, there are over XXX cluster in the internal regional of the galaxy, a region overwhich stream finder was not applied since the high stellar density implicates a computational cost that is currently far to expensive. Also, Balbinot demonstrates that internal dynamics preferentially kicks out small stars, which have lower luminosities and can make detection more difficult.

  One of the interests in studying stellar streams, is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of \paola{dark matter} haloes using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams that yield the most information about the galactic potential. However, the use of single streams for constraining the Galactic potential has led, however, to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, johnston05, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the radii of the Palomar~5 and GD~1 streams. Nonetheless, a recent goal of stellar stream astronomy was realized by \citet{2024ApJ...967...89I} constructed the most accurate model of the Milky Way to date based on a set of about 90 streams.

  In this study, we build on this body of work by investigating another important use case for stellar streams: probing dark matter on sub-galactic scales. According to simulations by Springel et al. (2008), $\Lambda$-Cold Dark Matter ($\Lambda$CDM) predicts that galaxies grow hierarchically, with dark matter clumps forming at a wide range of masses and sizes. These clumps, or \textit{subhalos}, are predicted to follow a mass distribution with a power-law slope slightly shallower than -2.0. To date, the smallest observed dark matter halo was detected through gravitational lensing in an Einstein ring, as described by \citet{2012Natur.481..341V}. However, some models predict that dark matter clumps could exist down to the mass of Earth-like planets at least \citep[see][ and discussion in \citet{2021arXiv211101148A}]{2005JCAP...08..003G, wang20}.

  \citet{2002MNRAS.332..915I} first suggested that dark matter subhalos could influence stellar streams by diffusing their orbital elements. Later, \citet{2012ApJ...748...20C} expanded this idea, proposing that subhalos could create gaps in stellar streams during flyby encounters, where a subhalo approaches closely enough to a segment of a stream and significantly changes the orbits of the closest stars. Carlberg 2013 further explored this idea through a probabilistic model of encounter rates based on the expected subhalo distribution. \citet{2019ApJ...880...38B} provided observational evidence for this idea, identifying an underdensity in the GD-1 stream that could not be explained by known objects, such as globular clusters, and, interestingly enough, was inconsistent with the $\Lambda$CDM mass-size relationship from M et al 20XX.

  Our study focuses on the influence of globular clusters on stellar streams, specifically examining whether globular clusters could produce gaps similar to those expected from dark matter subhalos. If globular clusters do indeed create such gaps, this factor must be considered when using stellar streams to detect dark matter. An additional and unique aspect of our study is that we focus on modeling real stellar streams and not just pure simulated streams. We focus on the Palomar 5 stream, which is well-studied, has observable gaps \citep{2020ApJ...889...70B}, and is highly visible due to its position above the galactic disk. We present a result from one of our simulations in Fig.~\ref{fig:stream_on_sky}, and the following sections describe our modeling of the stream, the gravitational interactions with the galaxy and other globular clusters, and the statistical analysis of perturbations affecting the stream.

@@

      \citet{2016MNRAS.457.3817S} extended the analysis of \citet{2015MNRAS.450.1136E}, demonstrating that action-angle variables provide a useful coordinate system for analyzing stream evolution, as actions are conserved and the angles associated with the stream's growth evolve linearly over time. Although we became aware of this work only after completing our analysis, we note that $\tau$ is a suitable approximation and behaves similarly to the angle variable corresponding to the azimuthal action: $\tau \approx \theta_{\phi,i} - \theta_{\phi,\text{GC}}$.
